---
title: "Lab8"
output: pdf_document
date: "2023-04-11"
---

```{r setup, include=FALSE}
library(tidyverse)
library(haven)
library(sandwich)
library(rdrobust)
library(rpart)
library(randomForest)
library(statar)

dat <- read_dta("health.dta")
view(dat)
```


```{r q1}

#Q1: Creating 10/90 training/test data set

#Set seed
HUID <- 21519588
set.seed(HUID)

#Store health variables from time t-1 which all end with _tm
all_predictors <- colnames(dat[,grep("^[tm1]", names(dat))])
all_predictors

#Store predictor variables which all start with P_*, but EXCLUDE race
race <- c("tm1_dem_black")
exclude_race <- setdiff(all_predictors,race)
exclude_race

#Define training and test data sets
#Use a uniformly distributed random number between 0 and 1
dat$random_number <- runif(length(dat$patient_id))

## Generate a training flag for 10% of the sample
dat$train_flag <- ifelse(dat$random_number<= 0.1, 1, 0) 

#Report number of observations in training and test samples
sum(dat$train_flag)
sum(1-dat$train_flag)

#Data frame with training data (randomly selected 10% of the data)
training <- subset(dat, train_flag == 1)
summary(training)

#Data frame with test data (remaining 90% of the data)
test <- subset(dat, train_flag == 0)
summary(test)



```



```{r q2}

#Q2: Estimating random forest models using the training data set

#2A: random forest model predicting patient costs, excluding race
mod1 <- randomForest(reformulate(exclude_race, "cost_t"), 
                     ntree=100, 
                     mtry=149,
                     importance=TRUE,
                     data=training)
mod1

### generate predictions for all observations in test and training samples
y_test_predictions_mod1 <- predict(mod1, newdata=test)
y_train_predictions_mod1 <- predict(mod1, newdata=training)

#Variable importance
importance(mod1)
varImpPlot(mod1, type=1) #Plot the Random Forest Results
dev.copy(png,'mod1_importance.png')
dev.off()

#2B: random forest model predicting patient costs, including race
mod2 <- randomForest(reformulate(all_predictors, "cost_t"), 
                     ntree=100, 
                     mtry=150,
                     importance=TRUE, 
                     data=training)
mod2

y_train_predictions_mod2 <- predict(mod2, newdata=training)
y_test_predictions_mod2 <- predict(mod2, newdata=test)

#Variable importance
importance(mod2)
varImpPlot(mod2, type=1) #Plot the Random Forest Results
dev.copy(png,'mod2_importance.png')
dev.off()


#2C: random forest model predicting patient health, excluding race
mod3 <- randomForest(reformulate(exclude_race, "gagne_sum_t"), 
                        ntree=100, 
                        mtry=149,
                        importance=TRUE, 
                        data=training)
mod3

y_test_predictions_mod3 <- predict(mod3, newdata=test)
y_train_predictions_mod3 <- predict(mod3, newdata=training)


#Variable importance
importance(mod3)
varImpPlot(mod3, type=1) #Plot the Random Forest Results
dev.copy(png,'mod3_importance.png')
dev.off()


#2D: random forest model predicting patient health, including race
mod4 <- randomForest(reformulate(all_predictors, "gagne_sum_t"), 
                     ntree=100, 
                     mtry=150,
                     importance=TRUE, 
                     data=training)
mod4

y_train_predictions_mod4 <- predict(mod4, newdata=training)
y_test_predictions_mod4 <- predict(mod4, newdata=test)

#Variable importance
importance(mod4)
varImpPlot(mod4, type=1) #Plot the Random Forest Results
dev.copy(png,'mod4_importance.png')
dev.off()






```
```{r q3}

#Q3: Comparing RMSPE for models including/excluding race in the training sample:
p <- 4
RMSPE <- matrix(0, p, 1)

## Model 1
RMSPE[1] <- sqrt(mean((training$cost_t - y_train_predictions_mod1)^2, na.rm=TRUE))

## Model 2
RMSPE[2] <- sqrt(mean((training$cost_t - y_train_predictions_mod2)^2, na.rm=TRUE))

## Model 3 
RMSPE[3] <- sqrt(mean((training$gagne_sum_t - y_train_predictions_mod3)^2, na.rm=TRUE))

## Model 4
RMSPE[4] <- sqrt(mean((training$gagne_sum_t - y_train_predictions_mod4)^2, na.rm=TRUE))

#Display a table of the results
data.frame(algorithm = c("Model 1 - Costs (excl. race) ", 
                             "Model 2 - Costs (incl. race) ",
                             "Model 3 - Health (excl. race)",
                             "Model 4 - Health (incl. race)"),
           RMSPE)


```

Using training data, between the models that predict patient costs, including race variables leads to a higher RMSPE. Between the models that predict patient health, including race variables leads to a lower RMSPE.

```{r q4}

#Q4: Comparing RMSPE for models including/excluding race in the test sample:

p <- 4
RMSPE_OOS <- matrix(0, p, 1)

## Model 1
RMSPE_OOS[1] <- sqrt(mean((test$cost_t - y_test_predictions_mod1)^2, na.rm=TRUE))

## Model 2
RMSPE_OOS[2] <- sqrt(mean((test$cost_t - y_test_predictions_mod2)^2, na.rm=TRUE))

## Model 3
RMSPE_OOS[3] <- sqrt(mean((test$gagne_sum_t - y_test_predictions_mod3)^2, na.rm=TRUE))

## Model 4
RMSPE_OOS[4] <- sqrt(mean((test$gagne_sum_t - y_test_predictions_mod4)^2, na.rm=TRUE))


#Display a table of the results
data.frame(algorithm = c("Model 1 - Costs (excl. race) ", 
                         "Model 2 - Costs (incl. race) ",
                         "Model 3 - Health (excl. race)",
                         "Model 4 - Health (incl. race)"),
           RMSPE_OOS)

```

Testing the models using the test data, between the models that predict patient costs, including race variables leads to a higher RMSPE. Between the models that predict patient health, including race variables leads to a marginally higher RMSPE.

```{r q5}

#Q5: Exporting test dataset and predictions
lab8 <- test

lab8$y_test_predictions_mod1 <- y_test_predictions_mod1
lab8$y_test_predictions_mod2 <- y_test_predictions_mod2
lab8$y_test_predictions_mod3 <- y_test_predictions_mod3
lab8$y_test_predictions_mod4 <- y_test_predictions_mod4

write_dta(lab8, "lab8_2023_results.dta")


```


```{r q6}

#Q6: Converting predictions in the test sample into percentile ranks 

#Read in prediction data
pred_dat <- read_dta("lab8_2023_results.dta")

#Creating percentile ranks for Model 1 predictions 
pred_dat <- pred_dat |> 
          mutate(riskscore_1 = rank(y_test_predictions_mod1))

max_rank_1 <- max(pred_dat$riskscore_1)

pred_dat$riskscore_rank_1 <- (pred_dat$riskscore_1/max_rank)*100

percentile_rank <- function(variable){
  r <- ifelse(is.na(variable), NA, rank(variable, ties.method = "average"))
  100*r/max(r, na.rm = TRUE)
}

pred_dat$riskscore_1 <- with(pred_dat, percentile_rank(y_test_predictions_mod1))
view(pred_dat)


#Percentile ranks for Model 2 predictions
pred_dat <- pred_dat |> 
          mutate(riskscore_2 = rank(y_test_predictions_mod2))

max_rank_2 <- max(pred_dat$riskscore_2)

pred_dat$riskscore_rank_2 <- (pred_dat$riskscore_2/max_rank)*100

percentile_rank <- function(variable){
  r <- ifelse(is.na(variable), NA, rank(variable, ties.method = "average"))
  100*r/max(r, na.rm = TRUE)
}

pred_dat$riskscore_2 <- with(pred_dat, percentile_rank(y_test_predictions_mod2))

#Percentile ranks for Model 3 predictions
pred_dat <- pred_dat |> 
          mutate(riskscore_3 = rank(y_test_predictions_mod3))

max_rank_3 <- max(pred_dat$riskscore_3)

pred_dat$riskscore_rank_3 <- (pred_dat$riskscore_3/max_rank)*100

percentile_rank <- function(variable){
  r <- ifelse(is.na(variable), NA, rank(variable, ties.method = "average"))
  100*r/max(r, na.rm = TRUE)
}

pred_dat$riskscore_3 <- with(pred_dat, percentile_rank(y_test_predictions_mod3))

#Percentile ranks for Model 4 predictions 
pred_dat <- pred_dat |> 
          mutate(riskscore_4 = rank(y_test_predictions_mod4))

max_rank_4 <- max(pred_dat$riskscore_4)

pred_dat$riskscore_rank_4 <- (pred_dat$riskscore_4/max_rank)*100

percentile_rank <- function(variable){
  r <- ifelse(is.na(variable), NA, rank(variable, ties.method = "average"))
  100*r/max(r, na.rm = TRUE)
}

pred_dat$riskscore_4 <- with(pred_dat, percentile_rank(y_test_predictions_mod4))
view(pred_dat)





```


```{r q7}

#Evaluating risk scores for extra resources, including race

#7a: Creating indicator variables for each model when risk score is > 55
pred_dat$highrisk_1 <- ifelse(pred_dat$riskscore_1 > 55, 1, 0)
pred_dat$highrisk_2 <- ifelse(pred_dat$riskscore_2 > 55, 1, 0)
pred_dat$highrisk_3 <- ifelse(pred_dat$riskscore_3 > 55, 1, 0)
pred_dat$highrisk_4 <- ifelse(pred_dat$riskscore_4 > 55, 1, 0)

#7B: Reporting fraction of Black patients that would be eligible for the program using each of the four models

pred_dat_black <- pred_dat |> 
                  filter(tm1_dem_black == 1) |> 
                  summarize(mean(highrisk_1),
                            mean(highrisk_2),
                            mean(highrisk_3),
                            mean(highrisk_4))
pred_dat_black

#7C: Reporting fraction of eligible patients that are Black using each of the four models
hrisk1 <- pred_dat |> 
         filter(highrisk_1 == 1) |> 
         summarize(mean(tm1_dem_black))

hrisk2 <- pred_dat |> 
         filter(highrisk_2 == 1) |> 
         summarize(mean(tm1_dem_black))

hrisk3 <- pred_dat |> 
         filter(highrisk_3 == 1) |> 
         summarize(mean(tm1_dem_black))

hrisk4 <- pred_dat |> 
         filter(highrisk_4 == 1) |> 
         summarize(mean(tm1_dem_black))

black_table <- c(hrisk1, hrisk2, hrisk3, hrisk4)
black_table

```
Over half of all Black patients would be eligible for the program using each of the four algorithms. However, a higher proportion of Black patients would be eligible using models 3 and 4, which predict the label for patient health, as opposed to models 1 and 2, which predict the label for patient costs. 
Conversely, a higher proportion of eligible patients in models 3 and 4 are Black, as opposed to models 1 and 2. 

```{r q8}

#Creating binned scatterplots of patient costs and patient health vs. risk score for White and Black patients separately:

#Bin scatter plot – connected dots
#Model 1 
graph1 <- ggplot(pred_dat, aes(x = riskscore_1 , y = cost_t, color = race)) +
 stat_binmean(n = 20, geom = "line") +
 stat_binmean(n = 20, geom = "point")
graph1


#Model 2
graph2 <- ggplot(pred_dat, aes(x = riskscore_2 , y = cost_t, color = race)) +
 stat_binmean(n = 20, geom = "line") +
 stat_binmean(n = 20, geom = "point")
graph2

#Model 3
graph3 <- ggplot(pred_dat, aes(x = riskscore_3 , y = gagne_sum_t, color = race)) +
 stat_binmean(n = 20, geom = "line") +
 stat_binmean(n = 20, geom = "point")
graph3


#Model 4 
graph4 <- ggplot(pred_dat, aes(x = riskscore_4 , y = gagne_sum_t, color = race)) +
 stat_binmean(n = 20, geom = "line") +
 stat_binmean(n = 20, geom = "point")
graph4




```

***Question 9***
The target parameter used to build the algorithm is prone to bias in a way that can underrepresent certain populations, such as Black patients, from being "seen" by the algorithm as entering a higher threshold for extra patient care or resources. I agree because as seen on the graphs, Black and white patients are roughly the same in patient costs and health outcomes, but models 1 and 2 underrepresent Black patients as opposed to models 3 and 4.


